One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-10-11 doc/genesetdp.tex
pair edited text

The significance of the inferences from NEA is assessed by null models that investigate the number of expected random links from the query gene set to the pathway. Previous efforts have settled for various parametric null distributions, where the parameters are estimated by Monte Carlo-simulations that permute the network definitions by methods with varying degrees of topological conservation.

% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics. For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}. 
% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
The methods that are used for network randomization, while computationaly efficient, are often mentally untractable as they include a big number of steps that are subject to complex constrains.
As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow contrains that essentially depends on all previous steps.
And so, while in the end we do come to a network that is randomized while preserving some desired topological features, the nested constrains make it very hard to describe the probability space to which this network belongs. 
We will then come to a null hypothesis for which rejection is uninformative, defeating the purpose the statistical test.

Users of NEA typically want to test if there is a functional association between a query set of genes and a pathway\cite{alexeyenko2012network}. We argue that such test is much easier implemented as tests of the query set, i.e. is the query set a randomly selected set of genes or is there an over representation of well-connected genes in the query set?

Hence, here we employ a different null model, more directly aimed at modeling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''. We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size, noting here that we do not need to make any assumptions on the nature of the score distribuition. We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
The algorithms enabled us to calculate well-defined statistics of inferences from NEA that we argue is more intuitive from an end-user perspective and thus more likely to be interpreted in a correct manner.
Using simulations we demonstrated that both the algorithms produced unbiased statistics.

>>>>

The significance of the inferences from NEA is assessed by null models that investigate the number of expected random links from the query gene set to the pathway. Previous efforts have settled for network randomization, where the network definitions are randomly constructed by methods with varying degrees of topological conservation in respect to the original network.

% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics. For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
The methods that are used for network randomization, while computationaly efficient, are often mentally untractable as they include a big number of steps that are subject to complex constrains.
As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow contrains that depends on previous link assignments.
While this results in a network that is both randomized and contain some preserved desired topological features, these nested constrains make it very hard to verify that the resulting network is representative. In theory there is a space of all possible networks that could be made up from a certain node degree distribution, however, it is not clear that the  second-order conservation method samples evenly from that distribution.  And if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose the statistical test.

Users of NEA typically want to test if there is a functional association between a query set of genes and a pathway\cite{alexeyenko2012network}. However, we argue that the network pertubation theory used in most flavors of NEA is a red herring. The typical user have used an experimental or computational method to derive a set of query genes that the user want to know if it is random, or if it is enriched in repect to a pathway.  In this question, the network is not relevant. A network may help to increase the sensitivity of ORA methods, but it is better left out of the null formation of the null hypothesis. We argue that such test is naturally implemented as tests of the query set, i.e. is the query set a randomly selected set of genes or is there an over representation of well-connected genes in the query set?

Hence, here we employ a different null model, more directly aimed at modeling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''. We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size, noting here that we do not need to make any assumptions on the nature of the score distribuition. We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
Combining the GeneSetDP and  GeneSetMC with the scoring system of the popular method BinoX, we using simulations to demonstrate that both our algorithms produced unbiased statistics.
